Quantifying the home ground advantage with `hyper2` and `hyper3` objects

### 2025-02-07 inst/home_advantage.Rmd
chi-square analysis

```{r,label=libload,echo=FALSE,print=FALSE,warning=FALSE,include=FALSE}
defaultW <- getOption("warn") 
options(warn = -1) 
library("hyper2",quietly=TRUE)
options(warn = defaultW)
```

@@

```{r showdata}
a <- read.table("celtic_rangers_livingston_falkirk.txt", header=TRUE)
head(a)
(aF <- home_away_table(a, give=FALSE))
(aT <- home_away_table(a, give=TRUE))
```

```{r doscottish,cache=TRUE}
(H <- home_away(aF))
(mH <- maxp(H))
```

@@

Now consider Celtic vs Falkirk, Celtic at home and also  Falkirk at home.

### (Celtic at home) {-}
* hyper2: P(Celtic;Falkirk) = $\frac{390+67; 56}{390 + 56 + 67}\simeq (0.89; 0.11)$
* hyper3: P(Celtic;Falkirk) = $\frac{1.77*412; 74}{1.77*412+74}\simeq(0.92; 0.08)$

@@

###  (Celtic away)
* hyper2: P(Celtic;Falkirk) = $\frac{390; 56+67}{390 + 56 + 67}\simeq (0.76;0.24)$
* hyper3: P(Celtic;Falkirk) = $\frac{412; 1.77*74}{412 + 1.77*74}\simeq(0.76; 24)$

@@

>>>>

```{r,label=libload,echo=FALSE,print=FALSE,warning=FALSE,include=FALSE}
defaultW <- getOption("warn") 
options(warn = -1) 
library("hyper2",quietly=TRUE)
library("magrittr")
options(warn = defaultW)
```

@@

```{r showdata}
a <- read.table("celtic_rangers_livingston_falkirk.txt", header=TRUE)
head(a)
teams <- c("Rangers","Celtic","Falkirk","Livingston")
(aF <- home_away_table(a, give=FALSE,teams = teams))
(aT <- home_away_table(a, give=TRUE, teams = teams))
```

(from here on we ignore draws and consider only won games)

```{r doscottish,cache=TRUE}
(H <- home_away(aF))
pnames(H) <- c(teams,"home")
H
(mH <- maxp(H))
```

@@

# Rangers vs Falkirk {-}

**(Rangers at home)**

* hyper2: P(Rangers;Falkirk) = $\frac{448+67; 56}{448 + 56 + 67}\simeq (0.902; 0.098)$
* hyper3: P(Rangers;Falkirk) = $\frac{1.77*448; 56}{1.77*448+56}\simeq(0.934; 0.066)$

actual scoreline: 69(R) / 2(F) or about $(0.972; 0.028)$. Expectation
would be (64;7) (H2) or (66.3;4.7) (H3)

*(Rangers away)*

* hyper2: P(Rangers;Falkirk) = $\frac{448; 56+67}{448 + 56 + 67}\simeq (0.78;0.22)$
* hyper3: P(Rangers;Falkirk) = $\frac{448; 1.77*56}{448 + 1.77*56}\simeq(0.82; 0.18)$

actual scoreline: 57(R) /17(F) - or about $(0.77; 0.23)$

# Rangers vs Livingston {-}

**(Rangers at home)**

* hyper2: P(Rangers;Livingston) = $\frac{448+67; 39}{448 + 39 + 67}\simeq (0.930; 0.070)$
* hyper3: P(Rangers;Livingston) = $\frac{1.77*448; 39}{1.77*448+39}\simeq(0.953; 0.047)$

actual scoreline:  15(R) / 0(L) or  $(1; 0)$

*(Rangers away)*

* hyper2: P(Rangers;Livingston) = $\frac{448; 39+67}{448 + 39 + 67}\simeq (0.81;0.19)$
* hyper3: P(Rangers;Livingston) = $\frac{448; 1.77*39}{448 + 1.77*39}\simeq(0.86; 0.14)$

actual scoreline: 12(R) /2(L) - or about $(0.86; 0.14)$

# Celtic vs Falkirk {-}

*(Celtic at home)*

* hyper2: P(Celtic;Falkirk) = $\frac{390+67; 56}{390 + 56 + 67}\simeq (0.89; 0.11)$
* hyper3: P(Celtic;Falkirk) = $\frac{1.77*412; 74}{1.77*412+74}\simeq(0.92; 0.08)$

@@

*(Celtic away)*

* hyper2: P(Celtic;Falkirk) = $\frac{390; 56+67}{390 + 56 + 67}\simeq (0.76;0.24)$
* hyper3: P(Celtic;Falkirk) = $\frac{412; 1.77*74}{412 + 1.77*74}\simeq(0.76; 0.24)$

@@

# Livingston vs Falkirk {-}

*(Livingston at home)*

* hyper2: P(Livingston;Falkirk) = $\frac{39+67; 56}{39 + 67 + 56}\simeq (0.65; 0.35)$
* hyper3: P(Livingston;Falkirk) = $\frac{1.77*42; 74}{1.77*42+74}\simeq(0.50; 0.50)$

actual scoreline: 3-9 or about $(0.25; 0.75)$

*(Livingston away)*

* hyper2: P(Livingston;Falkirk) = $\frac{39; 56+67}{39 + 56 + 67}\simeq (0.24;0.76)$
* hyper3: P(Livingston;Falkirk) = $\frac{42; 1.77*74}{42 + 1.77*74}\simeq(0.24; 0.76)$

actual scoreline: 6-8 or about $(0.43; 0.57)$

# Expectations

First, hyper2:

```{r}
mH
mH["Rangers"]
mH["home"]
mH3
lambda <- o$maximum
h2_exp <- matrix(0,4,4)
rownames(h2_exp) <- teams
colnames(h2_exp) <- teams
h3_exp <- h2_exp
n_matches <- Re(aF) + Im(aF)
for(hometeam in teams){
    for(awayteam in teams){

        h2_exp[hometeam,awayteam] <- (
            n_matches[hometeam,awayteam] * (mH[hometeam] + mH["home"])/
            (mH[hometeam] + mH[awayteam] + mH["home"])
        )
        h2_exp[hometeam,awayteam] %<>%
            add(1i*
                n_matches[hometeam,awayteam] * (mH[awayteam])/
                (mH[hometeam] + mH[awayteam] + mH["home"])
                )

        h3_exp[hometeam,awayteam] <- (
            n_matches[hometeam,awayteam] * (lambda* mH3[hometeam])/
            (lambda* mH3[hometeam] + mH3[awayteam])
        )

        h3_exp[hometeam,awayteam] %<>%
            add(1i*
                n_matches[hometeam,awayteam] * (mH3[awayteam])/
                (lambda* mH3[hometeam] + mH3[awayteam])
        )
    }
}
aF
round(h2_exp,1)
round(h3_exp,1)

makev <- function(M){
    out <- c(Re(M),Im(M))
    out[!is.na(out)]
}

(obs <- makev(aF))
(exp2 <- makev(h2_exp))
(exp3 <- makev(h3_exp))

plot((obs-exp2)^2/exp2)
plot((obs-exp3)^2/exp3)

chisq2 <- sum((obs-exp2)^2/exp2)
chisq3 <- sum((obs-exp3)^2/exp3)

pchisq(chisq2,df=12-4,lower.tail=FALSE)
pchisq(chisq3,df=12-4,lower.tail=FALSE)

par(pty='s')
plot(obs,exp2,log="xy",asp=1,xlim=c(1,100),ylim=c(1,100))
abline(0,1)
plot(obs,exp3,log="xy",asp=1,xlim=c(1,100),ylim=c(1,100))
abline(0,1)
```
